---
title: '2017 week 39: Food Fashions'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(stringr)
```

It's week 39 of [Makeover Monday](http://www.makeovermonday.co.uk/data/) 2017. This week focuses on food trends, with data from [nielsen](http://www.nielsen.com/content/dam/nielsenglobal/eu/docs/pdf/Global%20Ingredient%20and%20Out-of-Home%20Dining%20Trends%20Report%20FINAL%20(1).pdf).

```{r}
d <- readxl::read_excel("data/2017-week-39.xlsx")
```


```{r}
ggplot(d, aes(x = Region, y = Followers)) + 
    geom_col() + facet_wrap(~ reorder(Diet, Followers, sum)) + 
    gghlab::tilt_x_labels(90)
```

The most popular dietary fashions are for nutrient restriction: low carb/salt/sugar/fat. Europe in general seems to be lower than the others in most cases. The most dramatic difference across countries is in the Halal category. 

```{r}
ggplot(d, aes(x = reorder(Diet, Followers, sum), y = Followers, color = Region)) + 
    geom_point() + geom_line(aes(group = Region)) + 
    gghlab::tilt_x_labels(90)
```

This shows more clearly that Europe tends to be lowest in most categories, with North America typically following next. Asia-Pacific and Latin America are on the higher ends of things. 

```{r}
d <- d %>% 
    mutate(Group = case_when(
        str_detect(Diet, "Free|Low|Conscious") ~ "Nutrient-focused", 
        Diet %in% c("Kosher", "Halal") ~ "Religious",
        TRUE ~ "Low animal"
    ))

ggplot(d, aes(x = reorder(Diet, Followers, sum), y = Followers, 
        color = reorder(Region, -Followers))) + 
    geom_point() + geom_line(aes(group = Region)) + 
    gghlab::tilt_x_labels(90) + 
    facet_wrap(~ Group, scales = "free_x")
```

Dividing it into groups like this gives a little more context. 

```{r}
d %>% 
    group_by(Group,  Diet) %>% 
    mutate(Normalized = Followers - median(Followers)) %>% 
    ggplot(aes(x = reorder(Diet, Followers, sum), y = Normalized, 
        color = reorder(Region, -Followers))) + 
    geom_point() + geom_line(aes(group = Region)) + 
    gghlab::tilt_x_labels(90) + 
    facet_wrap(~ Group, scales = "free_x")
```

This shows it normalized, with the value for each regions/diet subtracted by the median value (across regions), so positive numbers mean more prominent than the median and negative mean less prominent. 

```{r}
# principal components
# used log10(Followers) bc not very normally distributed
# hist(d$Followers) vs hist(log10(d$Followers))
pca <- d %>% 
    select(-Group) %>% 
    mutate(Followers = log10(Followers)) %>% 
    spread(Diet, Followers) %>% 
    .[,2:12] %>% prcomp() %>% 
    .$x

pca <- cbind(pca, spread(select(d, -Group), Diet, Followers)[1:5, "Region"])

ggplot(pca, aes(x = PC1, PC2, color = Region)) + geom_point()
```

```{r, warn=FALSE}
d %>% 
    select(-Group) %>% 
    spread(Region, Followers) %>% 
    .[,2:6] %>% 
    cor() 

to_correlate <- d %>% 
    select(-Group) %>% 
    spread(Region, Followers) 

GGally::ggpairs(to_correlate[,2:6]) + ggtitle("with Halal")
GGally::ggpairs(to_correlate[to_correlate$Diet != "Halal",2:6]) + ggtitle("without Halal")
```

Weak correlation between Africa/Middle East with other regions is based on Halal diet outlier. If you remove that category, the correlations are very strong across all regions. 

```{r}
d %>% 
    group_by(Region) %>% 
    mutate(DietRank = 12 - rank(Followers, ties.method = "random")) %>% 
    select(Region, Diet, DietRank) %>% 
    ggplot(aes(x = reorder(Diet, DietRank, median), y = DietRank, 
        color = Region, group = Region)) + 
    geom_point() + gghlab::tilt_x_labels() + geom_line()

```


Final version: 
```{r}
to_plot <- d %>% 
    mutate(Diet = case_when(
        str_detect(Diet, "Wheat") ~ "Low gluten", 
        str_detect(Diet, "Lactose") ~ "Low dairy",
        str_detect(Diet, "Sugar") ~ "Low sugar",
        str_detect(Diet, "Carb") ~ "Low carb",
        str_detect(Diet, "Fat") ~ "Low fat",
        str_detect(Diet, "Sodium") ~ "Low salt",
        TRUE ~ Diet
    )) %>% 
    mutate(Region = ifelse(Region == "Africa/Middle East", 
        "Africa/Mid-East", Region))

ggplot(to_plot, aes(x = reorder(Diet, Followers, sum), 
        y = Followers * 100, color = Region)) + 
    geom_line(aes(group = Region)) + geom_point() + 
    gghlab::tilt_x_labels() + 
    facet_grid(.~Group, scales = "free_x", space = "free_x") + 
    ylab("% of people in region following diet") + 
    xlab(NULL) + 
    scale_color_brewer(palette = "Set1") + 
    labs(color = NULL, title = "Diets are similar around the world", 
        subtitle = paste(
            "Diets popular in one region tend to be popular in others.", 
            "The biggest regional difference,\nother than halal cuisine, is that", 
            "Europeans report the least restrictive eating patterns."))

ggsave("out/2017-week-39.png", width = 6, height = 4, units = "in", dpi = 303)

```
