---
title: '2017 week 33: Births in the US'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(stringr)
library(dplyr)
```

It's week 33 of [Makeover Monday](http://www.makeovermonday.co.uk/data/) 2017. The data is about the [timing of births in America](https://pudding.cool/2017/05/births/index.html) and comes from the [CDC](https://www.cdc.gov/nchs/data_access/vitalstatsonline.htm). 

There are so many interesting variables in this data set! You can look at geographic and temporal variation of

* mother's age
* child's gender
* child's birth weight
* number of births

```{r}
# there is a weird byte at the beginning of the file which messes up the column names
d <- read.table("data/2017-week-33.csv", sep = "\t", skipNul = TRUE, skip = 1)

colnames(d) <- readLines("data/2017-week-33.csv", n = 1, skipNul = TRUE) %>% 
    str_sub(3, -1) %>% 
    str_split("\t") %>% 
    .[[1]] %>% 
    str_replace_all(" ", "")

d <- d %>% 
    as_tibble() %>% 
    mutate(Date = lubridate::dmy(Date))

state_abbrevs <- read.csv("data/2017-week-33-states.csv")

d <- inner_join(d, state_abbrevs)
```


```{r}
ggplot(d, aes(x = Births, color = factor(MonthCode))) + 
    geom_freqpoly(alpha = 0.5) + 
    scale_x_log10()

ggplot(d, aes(x = Births, color = factor(Year))) + 
    geom_freqpoly() + scale_x_log10()

ggplot(d, aes(y = Births, x = factor(MonthCode))) + 
    geom_boxplot()
```

What's up with the weird discontinuity where births are either above or below 4000? 

```{r}
filter(d, Births > 4000) 
```

It turns out to be the data for Los Angeles County, which is apparently unique across the nation for having dramatically more births than any other county. 

```{r}
ggplot(d, aes(y = Births, x = factor(MonthCode))) + 
    geom_boxplot() + scale_y_log10()
```

## Mother's age
```{r}
ggplot(d, aes(y = AverageAgeofMother, x = Year, group = Year)) + 
    geom_boxplot() 

ggplot(d, aes(y = AverageAgeofMother, x = Year)) + 
    stat_smooth(method = lm) + 
    facet_wrap(~ reorder(State, AverageAgeofMother, mean)) + 
    ggtitle("by mean age of mother")

x <- d %>%
    group_by(State) %>%
    group_by(State, Year) %>%
    summarize(MotherAge = weighted.mean(AverageAgeofMother, Births)) %>% 
    ungroup()

filter(x, Year == min(Year) | Year == max(Year)) %>% 
    ggplot(aes(x = Year, y = MotherAge, group = State)) + geom_line() + 
        geom_text(data = filter(x, Year == 2015), aes(label = State), hjust = 0) + 
        xlim(c(2002, 2018))

ggplot(x, aes(y = MotherAge, x = Year)) + 
    geom_line(aes(group = State), color = "grey50") + 
    geom_text(data = filter(x, Year == 2015), aes(label = State), hjust = 0) + 
    xlim(c(2002, 2018))


filter(x, Year == min(Year) | Year == max(Year)) %>% 
    ggplot(aes(y = reorder(State, MotherAge, min), x = MotherAge)) + 
        geom_line(aes(group = State), color = "grey50") + 
        geom_point(aes(fill = factor(Year)), pch = 21) + 
        labs(y = NULL, x = "Average mother's age at delivery", color = NULL) + 
        ggtitle("Women are giving birth later") + 
        scale_fill_manual(values = c("white", "black")) + 
        theme(legend.position = c(0.875, 0.05)) + 
        theme(legend.title = element_blank(),
            legend.margin = margin(0, 5, 5, 5))

ggsave("out/2017-week-33-a.png", height = 7, width = 4, units = "in", dpi = 303)

ggplot(d, aes(y = AverageAgeofMother, x = Year)) + 
    stat_smooth(method = lm, aes(group = State)) +
    scale_color_discrete(guide = FALSE) + 
    ggtitle("linear model of trend")

county_ages <- d %>%
    group_by(County) %>% 
    filter(Year == min(Year)) %>% 
    summarize(MotherStartAge = mean(AverageAgeofMother))

county_models <- d %>% 
    group_by(County) %>% 
    do(broom::tidy(lm(AverageAgeofMother ~ Year, data = .)))

counties <- inner_join(county_ages, county_models)

filter(counties, term == "Year") %>% 
    ggplot(aes(x = MotherStartAge, y = estimate)) + 
    geom_point() + 
    geom_hline(yintercept = 0) +
    ylab("Increase in mother's age per year") + 
    xlab("Average mother's age in 2003") + 
    ggtitle("County level linear models of mother's age vs year")

State_ages <- d %>%
    group_by(State) %>% 
    filter(Year == min(Year)) %>% 
    summarize(MotherStartAge = weighted.mean(AverageAgeofMother, Births))

State_models <- d %>% 
    group_by(State, Abbreviation) %>% 
    do(broom::tidy(lm(AverageAgeofMother ~ Year, data = .)))

states <- inner_join(State_ages, State_models)

filter(states, term == "Year") %>% 
    ggplot(aes(x = MotherStartAge, y = estimate)) + 
    ggrepel::geom_text_repel(aes(label = Abbreviation), 
        size = 2, min.segment.length = unit(0.1, "lines")) + 
    geom_point(color = "grey50") + 
    geom_hline(yintercept = 0) +
    ylab("Increase in mother's age per year\n(2003-2015)") + 
    xlab("Average mother's age in 2003") + 
    ggtitle("Women are giving birth later") + 
    labs(subtitle = 
        paste0("Since 2003, the average age at delivery has gone up in every state, ", 
            "with\nfaster rises in states where mothers were younger in 2003 ", 
            "(typically the\nsouth and southwest) and slower rises in states ", 
            "where mothers were\nalready giving birth at almost 30 in 2003."))

ggsave("out/2017-week-33-b.png", width = 5.25, height = 4, units = "in", dpi = 303)
```

## Birth weight
```{r}
ggplot(d, aes(y = AverageBirthWeight, x = AverageAgeofMother)) + 
    geom_hex() + 
    stat_smooth(method = lm) 

ggplot(d, aes(y = AverageBirthWeight, x = AverageAgeofMother)) + 
    geom_hex() + 
    stat_smooth(method = lm) + 
    facet_wrap(~ Year)

state_weight <- d %>%
    group_by(State, Year, Gender, Abbreviation) %>% 
    summarize(Weight = weighted.mean(AverageBirthWeight, Births),
        AverageAgeofMother = weighted.mean(AverageAgeofMother, Births),
        Births = sum(Births))

ggplot(state_weight, aes(x = Year, y = Weight)) + 
    geom_point() + geom_line(aes(group = State)) + 
    facet_wrap(~Gender) + 
    ggtitle("On average, male babies are bigger")

ggplot(d, aes(x = AverageBirthWeight, color = Gender, fill = Gender)) +
    geom_freqpoly(bins = 50, aes(weight = Births), stat = "bin") + 
    ggtitle("On average, male babies are bigger") 

lm(AverageBirthWeight ~ AverageAgeofMother + State + Year + Gender, data = d) %>% summary()

to_plot <- state_weight %>% 
    filter(Year == 2015) %>% mutate(Weight = Weight / 1000)
highlight <- to_plot %>% 
    filter(Weight > 3.45 | Weight < 3.1 | (AverageAgeofMother > 30 & Weight > 3.35) | 
        (AverageAgeofMother > 29 & Weight < 3.15) | (AverageAgeofMother < 27 & Weight > 3.32)
        )
ggplot(to_plot, aes(x = AverageAgeofMother, y = Weight)) + 
    geom_line(aes(group = State), color = "grey75") + 
    ggrepel::geom_text_repel(data = highlight, aes(label = Abbreviation), size = 3) + 
    geom_point(aes(color = Gender)) + 
    scale_color_manual(values = c("#ff748c", "#4b93c3"), guide = FALSE) + 
    scale_y_continuous(sec.axis = sec_axis(~ . * 2.20462, 
        name = "Average birth weight in 2015 (lb)")) + 
    stat_smooth(method = lm, aes(group = Gender, color = Gender), se = FALSE) + 
    ggtitle("Birth weight determined by gender, mother's age, and state") + 
    labs(y = "Average birth weight in 2015 (kg)", x = "Mother's age at delivery", subtitle = 
        paste("Average birth weights vary by almost a pound. The most important factors are:\n",
            "  - Gender: boys (blue) weigh 110g (4oz) more than girls (pink)\n",
            "  - Mother's age: each year older adds 12g (0.5oz) to birth weight\n", 
            "  - State: each state (connected by lines) has its own base weight"))
ggsave("out/2017-week-33-c.png", width = 5.5, height = 5, units = "in", dpi = 303)
```

## Birth seasonality
```{r}
ggplot(d, aes(y = Births, x = Date, group = County)) + 
    geom_line() + scale_y_log10() + facet_wrap(~Gender, ncol = 1) + 
    ggtitle("County")

d %>% 
    group_by(State, Date, Gender) %>% 
    summarise(Births = sum(Births)) %>% 
    ggplot(aes(y = Births, x = Date, group = State)) + 
    geom_line() + scale_y_log10() + facet_wrap(~Gender, ncol = 1) + 
    ggtitle("State")

d %>% 
    # sum across counties
    group_by(State, MonthCode, Gender, Year) %>% 
    summarise(Births = sum(Births)) %>% 
    # average across years
    group_by(State, MonthCode, Gender) %>% 
    summarise(Births = mean(Births)) %>% 
    ggplot(aes(y = Births, x = MonthCode, group = State)) + 
    geom_line() + scale_y_log10() + facet_wrap(~Gender, ncol = 1) + 
    ggtitle("State by month")


d %>% 
    group_by(State, Date, Gender) %>% 
    summarise(Births = sum(Births)) %>% 
    filter(State == "Washington") %>% 
    ggplot(aes(y = Births, x = Date, group = State, color = Gender)) + 
    geom_line() + 
    ggtitle("Washington")

d %>% 
    filter(State == "Washington") %>% 
    # sum across counties
    group_by(State, MonthCode, Gender, Year) %>% 
    summarise(Births = sum(Births)) %>% 
    # average across years
    group_by(State, MonthCode, Gender) %>% 
    summarise(Births = mean(Births)) %>% 
    ggplot(aes(y = Births, x = MonthCode, group = State)) + 
    geom_line() + facet_wrap(~Gender, ncol = 1) + 
    ggtitle("Washington by month")
```

## Gender
```{r}
d %>% 
    select(Date, County, Gender, Month, State, Year, Abbreviation, Births) %>% 
    tidyr::spread(Gender, Births) %>% 
    mutate(BoysPerGirl = Male / Female) %>% 
    ggplot(aes(x = Male + Female, y = BoysPerGirl)) + 
        geom_point(alpha = 0.1) + 
        geom_hline(yintercept = 1) + 
        scale_x_log10() + 
        stat_smooth()

```