---
title: '2017 week 30: Water footprints'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggrepel)
```

It's week 30 of [Makeover Monday](http://www.makeovermonday.co.uk/data/) 2017 edition. The data comes from [UNESCO](http://waterfootprint.org/media/downloads/Report-48-WaterFootprint-AnimalProducts-Vol1.pdf), and was covered [here](https://www.statista.com/chart/9483/how-thirsty-is-our-food/). 

```{r}
d <- readxl::read_excel("data/2017-week-30.xlsx")
d <- d[ , 1:8]
colnames(d) <- str_replace_all(colnames(d), " .*", "")

d <- d %>%
    tidyr::gather(Water, MeterPerTon, Blue, Green, Grey, Total) %>% 
    tidyr::gather(Nutrient, Amount, Fat, Protein, Calorie)
```

## Initial data exploration

```{r}
ggplot(d, aes(x = Water, y = MeterPerTon, fill = Food)) + 
    geom_col()

ggplot(d, aes(x = Water, y = MeterPerTon, fill = Food)) + 
    geom_col(position = "fill")
```

Most of the water used is green. There are too many groups to tell much about the groups. 

```{r}
food_categories <- data.frame(Food = c("Sugar crops", "Vegetables", 
    "Starchy roots", "Fruits", "Cereals", "Oil crops", "Pulses", 
    "Nuts", "Milk", "Eggs", "Chicken meat", "Butter", "Pig meat", 
    "Sheep/goat meat", "Bovine meat"), 
    Category = c("Crop", "Fruit/Veg", "Grain", "Fruit/Veg", "Grain", "Crop", 
        "Grain", "Nuts", "Dairy", "Dairy", "Meat", "Dairy", rep("Meat", 3)))
d <- inner_join(d, food_categories)

d %>% 
    group_by(Category, Water) %>% 
    summarise(MeterPerTon = mean(MeterPerTon)) %>% 
    ggplot(aes(x = Water, y = MeterPerTon, fill = Category)) + 
        geom_col(position = "fill") + 
        scale_fill_brewer(palette = "Set1")
```

Total water consumption from highest to smallest (per ton of product produced): meat, nut, dairy, grain, crop, and then fruit/veg. (average of all foods in the category)

```{r}
d %>% 
    filter(Amount > 0) %>% 
    group_by(Category, Water, Nutrient) %>% 
    summarise(MeterPerTon = mean(MeterPerTon), Amount = mean(Amount)) %>% 
    ggplot(aes(x = Water, y = MeterPerTon / Amount, fill = Category)) + 
    geom_col(position = "fill") + 
    scale_fill_brewer(palette = "Set1") + 
    facet_wrap(~ Nutrient)
```

Total water consumption per unit calorie/fat/protein is harder to interpret. Some foods 0 units of particular nutrients (e.g. butter has 0 g protein), so those foods are removed from this plot. 

Maybe a better way to think about it is: how many units of nutrient can I get per cubic meter of water? 

```{r}
d %>% 
    ggplot(aes(x = Water, y = Amount / MeterPerTon, color = Food)) + 
    geom_point() + facet_grid(Nutrient ~ Category, scales = "free_y")
```

Too much noise with some foods having very little use of one particular kind of water. 

```{r}
d %>% 
    filter(Water == "Total") %>% 
    ggplot(aes(x = Category, y = Amount / MeterPerTon, color = Food)) + 
    geom_point() + facet_wrap(~ Nutrient, scales = "free_y", ncol = 1)
```

Colors overwhelming but this is okay otherwise. Incidentally, I realized that the last three columns of the data are liter of water per unit nutrient (g fat/g protein/kcal), which is the inverse of the number I just came up with. I could have just used those columns, except that for cases where there is 0 of that unit, the numbers are off. 

```{r}

to_plot <- d %>% 
    mutate(UnitPerLiter = Amount / MeterPerTon) %>%
    mutate(Nutrient = case_when(
        Nutrient == "Fat" ~ "grams fat / liter water", 
        Nutrient == "Protein" ~ "grams protein / liter water",
        TRUE ~ "calories / liter water"
    )) %>% 
    filter(Water == "Total") %>% 
    group_by(Nutrient) %>%
    mutate(Highlight = UnitPerLiter > max(UnitPerLiter) * 0.75) %>% 
    mutate(Highlight = ifelse(Highlight, Food, NA))

ggplot(to_plot, aes(x = Category, y = UnitPerLiter)) + 
    geom_point(pch = 1) + 
    facet_wrap(~ Nutrient, scales = "free_y", ncol = 1) + 
    labs(y = NULL, x = NULL) + 
    ggrepel::geom_label_repel(aes(label = Highlight), 
        min.segment.length = unit(0.25, "lines"), max.iter = 10000) + 
    ggtitle("How many nutrients can a liter of water produce?") + 
    labs(subtitle = paste0("A liter of water produces the most calories and protein",
        " when converted\ninto grain. For fat, butter is the most efficient."))

ggsave("out/2017-week-30.png", width = 5, height = 5, units = "in", dpi = 303)
```
