---
title: '2017 week 27: German tourism'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggrepel)
```
It's week 27 of [Makeover Monday](http://www.makeovermonday.co.uk/data/) 2017 edition. The data this week is about tourism in Germany and comes from [statistik Berlin Brandenburg](https://www.statistik-berlin-brandenburg.de/grundlagen/tourismus-grafik.asp?Ptyp=260&Sageb=45000&creg=BBB&anzwer=5). 
```{r}
d <- readxl::read_excel("data/2017-week-27.xlsx")

colnames(d)[c(2, 6)] <- c("Destination", "From") 

# a few Nights values are NA, remove them so as not to throw off nights per guest
d <- d %>% 
    filter(complete.cases(.))

# summary by guest's home country
country_of_origin_summary <- d %>% 
    group_by(From) %>% 
    summarise(Guests = sum(Guests), Nights = sum(Nights), 
        NightsPerGuest = Nights / Guests) %>% 
    arrange(desc(Guests))
```

## summary by country of origin
```{r}
p <- function(d, f = head, ...) {
    d %>% 
    f(n = 8) %>% 
    ggplot(aes_string(...)) + 
        geom_point() +  
        gghlab::tilt_x_labels()
}

p_all <- function(d, ...) {
    d2 <- d %>% 
        mutate(Rank = 1:n())
    
    ggplot(d2, aes_string(...)) + 
        geom_point() +  
        theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
        geom_label_repel(data = filter(d2, Rank < 4 | Rank > (n() - 3)), 
            aes_string(...))
}

country_of_origin_summary %>% 
    arrange(desc(Guests)) %>% 
    p(x = "reorder(From, Guests)", y = "log10(Guests)") + ylim(c(5, 9)) + 
    ggtitle("Countries supplying most guests")

country_of_origin_summary %>% 
    arrange(NightsPerGuest) %>% 
    p(x = "reorder(From, NightsPerGuest)", y = "NightsPerGuest") +
    ggtitle("Countries whose guests stay shortest")

country_of_origin_summary %>% 
    arrange(desc(NightsPerGuest)) %>% 
    p(x = "reorder(From, NightsPerGuest)", y = "NightsPerGuest") +
    ggtitle("Countries whose guests stay longest")
```

I started out just looking at the top and bottom few. Then I started showing all the data and just labeling the top and bottom few, which is way better.

```{r}
country_of_origin_summary %>% 
    p_all(x = "reorder(From, Guests)", y = "log10(Guests)", label = "From") + 
    ggtitle("Number of guests per country")

country_of_origin_summary %>% 
    arrange(NightsPerGuest) %>% 
    p_all(x = "reorder(From, NightsPerGuest)", y = "NightsPerGuest", label = "From") + 
    ggtitle("Length of stay per country")
```

## summary by district visiting
```{r}
district_visited_summary <- d %>% 
    group_by(Destination) %>% 
    summarise(Guests = sum(Guests), Nights = sum(Nights), 
        NightsPerGuest = Nights / Guests)

district_visited_summary %>% 
    arrange(desc(Guests)) %>% 
    p_all(x = "reorder(Destination, Guests)", y = "log10(Guests)", label = "Destination") + 
    ggtitle("Number of visitors to destination in Germany")

district_visited_summary %>% 
    arrange(NightsPerGuest) %>% 
    p_all(x = "reorder(Destination, NightsPerGuest)", y = "NightsPerGuest", 
        label = "Destination") + 
    ggtitle("Length of stay in German destinations")

```

## Seasonal variation
```{r}
seasons <- d %>% 
    mutate(Year = factor(Year)) %>% 
    group_by(Mo, Year) %>% 
    summarise(Guests = sum(Guests), Nights = sum(Nights), 
        NightsPerGuest = Nights / Guests) 

ggplot(seasons, aes(x = Mo, y = Guests, color = Year)) + 
    geom_line() +
    ggtitle("Peak guests in spring and summer")

ggplot(seasons, aes(x = Mo, y = NightsPerGuest, color = Year)) + 
    geom_line() + 
    ggtitle("Peak nights per guest concentrated in summer")
```

## where do biggest visitors go
```{r}
longest <- c("Arabische Golfstaaten", "Israel", "Sonstige afrikanische Länder")

d %>% 
    filter(From %in% longest) %>% 
    group_by(From, Destination) %>% 
    summarise(NightsPerGuest = sum(Nights) / sum(Guests)) %>% 
    group_by(From) %>% 
    arrange(NightsPerGuest, .by_group = TRUE) %>% 
    mutate(rankx = 1:n()) %>% 
    p_all(x = "rankx", y = "NightsPerGuest", 
        label = "Destination") + 
        facet_wrap(~From, ncol = 2)



d %>% filter(From %in% "Arabische Golfstaaten") %>% 
    mutate(Year = factor(Year)) %>% 
    group_by(Destination, Year) %>% 
    summarise(g = sum(Guests), n = sum(Nights), npg = n/g) %>% 
    ggplot(aes(x = reorder(Destination, npg, mean), y = npg, color = Year)) + 
    geom_point()


d %>% filter(From %in% "Arabische Golfstaaten", 
    Destination %in% c("Märkisch-Oderland", "Ostprignitz-Ruppin", "Prignitz")) %>% 
    mutate(Year = factor(Year)) %>% 
    group_by(Destination, Year) %>% 
    summarise(g = sum(Guests), n = sum(Nights), npg = n/g) %>% 
    ggplot(aes(x = g, y = npg, color = Year)) + 
    geom_point() + facet_wrap(~Destination)


```

## where and when to go for most/fewest other tourists? 
```{r}
alone <- d %>% 
    mutate(Season = case_when(
        Mo > 2 & Mo < 6 ~ "Spring", 
        Mo > 5 & Mo < 9 ~ "Summer", 
        Mo > 8 & Mo < 12 ~ "Fall", 
        TRUE ~ "Winter")) %>% 
    group_by(Destination, Season) %>% 
    summarise(Guests = sum(Guests), NightsPerGuest = sum(Nights) / sum(Guests)) %>% 
    mutate(lab = ifelse((Guests < 1.45E5 & Season == "Winter"), Destination, ""), 
        lab = ifelse(lab == "" & Guests > 3E6 & Season == "Summer", Destination, lab))

ggplot(alone, aes(x = reorder(Destination, Guests, mean), y = Guests / 1E6)) + 
    geom_point(aes(color = Season)) + scale_y_log10() + 
    geom_label_repel(aes(label = lab), max.iter = 10E3, 
        point.padding = unit(0.1, "lines"), size = 3.5) + 
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) + 
    ggtitle("German destinations for introverts and extroverts") +
    labs(color = NULL, y = "Millions of visitors", x = "Destination")

ggsave("out/2017-week-27.png", width = 5, height = 2.5, units = "in", dpi = 303)
```