---
title: '2017 week 28: Tour de France'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggrepel)
```

```{r}
d <- readxl::read_excel("data/2017-week-28.xlsx")
colnames(d) <- colnames(d) %>% str_replace("'", "") %>%
    str_replace("\\(.*", "") %>% snakecase::to_big_camel_case()

```

# Change over time
```{r}
time_plot <- function(...) {
    ggplot(d, aes_string(x = "Year", ...)) + 
        geom_line() + 
        geom_point(aes(color = Winner == "Results voided")) + 
        labs(color = "Void?")
}

time_plot(y = "WinnersAvgSpeed") + ylim(c(0, 45))
time_plot(y = "NumberOfStages") + ylim(c(0, 26))
time_plot(y = "Finishers") + ylim(c(0, 210))
time_plot(y = "Entrants") + ylim(c(0, 210))
```

# Country and team summaries
```{r}
ggplot(d, aes(x = WinnersTeam)) + geom_bar() + gghlab::tilt_x_labels()
ggplot(d, aes(x = WinnersNationality)) + geom_bar() + 
    gghlab::tilt_x_labels()

d %>% 
    group_by(WinnersNationality) %>% 
    arrange(Year) %>% 
    mutate(Record = 1:n() / (Year - 1902)) %>% 
    ggplot(aes(x = Year, y = Record, color = WinnersNationality)) + 
    geom_line()

d %>% 
    group_by(WinnersNationality) %>% 
    arrange(Year) %>% 
    filter(n() > 1) %>% 
    mutate(Record = 1:n() / (Year - 1902)) %>% 
    ggplot(aes(x = Year, y = Record, color = WinnersNationality)) + 
    geom_line() + ggtitle("Countries with > 1 winner")

d %>% 
    group_by(WinnersNationality) %>% 
    arrange(Year) %>% 
    filter(n() > 2) %>% 
    mutate(Record = 1:n() / (Year - 1902)) %>% 
    ggplot(aes(x = Year, y = Record, color = WinnersNationality)) + 
    geom_line() + ggtitle("Countries with > 2 winners")

d %>% 
    group_by(WinnersNationality) %>% 
    summarise(LastWin = max(Year), Number = n()) %>% 
    ggplot(aes(x = LastWin, y = Number, color = WinnersNationality)) + 
    geom_point()

d %>% 
    group_by(WinnersNationality) %>% 
    summarise(LastWin = max(Year), Number = n()) %>% 
    arrange(desc(Number))
```

```{r}
best <- c("France", "Belgium", "Spain", "Italy")

working <- d %>% 
    mutate(WinnersNationality = str_trim(WinnersNationality)) %>%  
    filter(WinnersNationality != "Results voided") %>% 
    mutate(Country = ifelse(WinnersNationality %in% best,
        WinnersNationality, "Other"), Result = "Win") %>% 
    select(Country, Year, Result)

full <- data_frame(
    Year = rep(seq(min(d$Year), max(d$Year)), 5), 
    Country = rep(unique(working$Country), each = 114))

combined <- left_join(full, working, fill = NA) %>% 
    mutate(Result = ifelse(is.na(Result), "Loss", Result))

combined %>%    
    group_by(Country) %>% 
    arrange(Year) %>% 
    mutate(Record = cumsum(Result == "Win") / (Year - 1902)) %>% 
    ggplot(aes(x = Year, y = Record, color = Country)) + 
    geom_line()

to_plot <- combined %>%    
    group_by(Country) %>% 
    arrange(Year) %>% 
    mutate(Wins = cumsum(Result == "Win")) %>% 
    ungroup() %>% 
    mutate(Country = factor(Country, levels = 
            c("France", "Other", "Belgium", "Spain", "Italy")))

labels <- filter(to_plot, Year == max(Year)) %>% 
    mutate(Year = Year + 1)

ggplot(to_plot, aes(x = Year, y = Wins, color = Country)) + 
    geom_line() + 
    ggtitle("Increased diversity of Tour de France winners over time") + 
    labs(subtitle = "Early dominance of France and Belgium has waned since the 80s.", x = NULL, y = "Number of wins") + 
    scale_color_brewer(type = "qual", palette = "Set1", guide = FALSE) + 
    geom_text(data = labels, aes(label = Country), hjust = 0) + 
    xlim(c(1900, 2025))

ggsave("out/2017-week-28.png", width = 6, height = 4, units = "in", dpi = 303)
```

I like that graph a lot. I also tried drawing steps instead of lines, because wins are discrete events. This has its merits but I think the zig-zag pattern is distracting. 

```{r}
ggplot(to_plot, aes(x = Year, y = Wins, color = Country)) + 
    geom_step() + 
    ggtitle("Increased diversity of Tour de France winners over time") + 
    labs(subtitle = "Early dominance of France and Belgium has waned since the 80s.", x = NULL, y = "Number of wins") + 
    scale_color_brewer(type = "qual", palette = "Set1", guide = FALSE) + 
    geom_text(data = labels, aes(label = Country), hjust = 0) + 
    xlim(c(1900, 2025))
```

Then I wondered if representing the wins with points would more effectively show the changing win patterns. 

```{r}
filter(to_plot, Result == "Win") %>% 
    mutate(Country = factor(Country, levels = 
        rev(c("France", "Belgium", "Italy", "Spain", "Other")))) %>% 
    ggplot(aes(x = Year, y = Country)) + 
        geom_point(pch = "|", size = 4) + 
        scale_color_discrete(guide = FALSE) + 
        ggtitle("Increased diversity of Tour de France winners over time") + 
    labs(x = NULL, y = NULL, subtitle = 
            "Early dominance of France and Belgium has waned since the 80s.")

ggsave("out/2017-week-28-2.png", width = 5.25, height = 2.5, units = "in", dpi = 303)

```

I think this is a pretty effective way of showing it too. 