---
title: '2017 week 29: White House salaries'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(stringr)
library(dplyr)
library(ggrepel)
```

It's week 29 of [Makeover Monday](http://www.makeovermonday.co.uk/data/) 2017 edition. This week the data is a comparison of White House employee salaries between the Obama and Trump administrations. The data comes from [NPR](https://docs.google.com/spreadsheets/d/119OjfVU9Uf0lqfwu8VCjKsD73dMb5owmai5Nec1OFBI/edit#gid=0), which wrote about the topic [here](http://www.npr.org/2017/06/30/535069910/trump-white-house-staff-payroll-nearly-36-million-and-top-heavy). 

I struggled a lot with this one. 

```{r}
d <- bind_rows(
    readxl::read_excel("data/2017-week-29-a.xlsx", sheet = "Trump Admin"), 
    readxl::read_excel("data/2017-week-29-a.xlsx", sheet = "Obama admin"))
```

## Initial data exploration

Obama had more employees. Trump has fewer, who make, on average, more than the Obama employees, presumably because he has filled mainly higher level positions.
```{r}
d %>% 
    group_by(ADMINISTRATION) %>% 
    summarise(n = n(), mean = mean(SALARY), median = median(SALARY))
```

There are a ton of unique position titles. Can we put them in groups?

```{r}

pos <- function(pattern) {
    str_detect(d$`POSITION TITLE`, pattern)
}

d$Bin <- case_when(pos("^ADVISOR") ~ "Advisor", 
    pos("^ASSISTANT DIRECTOR") ~ "Assistant director",
    pos("^ASSISTANT TO THE PRESIDENT") ~ "Assistant to the President",
    pos("^ASSOCIATE COUNSEL") ~ "Associate counsel",
    pos("^ASSOCIATE DIRECTOR") ~ "Associate director",
    pos("^DEPUTY ASSISTANT") ~ "Deputy assistant",
    pos("^DEPUTY DIRECTOR") ~ "Deputy director",
    pos("^DIRECTOR") ~ "Director",
    pos("^EXECUTIVE ASSISTANT") ~ "Executive assistant", 
    pos("^INFORMATION SERVICES OPERATOR") ~ "Information services", 
    pos("^STAFF ASSISTANT") ~ "Staff assistant",
    pos("^SPECIAL ASSISTANT") ~ "Special assistant", 
    pos("^RECORDS MANAGEMENT") ~ "Records management",
    pos("^SENIOR POLICY") ~ "Senior policy advisor",
    TRUE ~ "Other")

d$Level <- case_when(pos("^ASSISTANT") ~ "Assistant",
    pos("^ASSOCIATE") ~ "Associate",
    pos("^CHIEF") ~ "Chief",
    pos("^DEPUTY") ~ "Deputy",
    pos("^DIRECTOR") ~ "Director",
    pos("^ASSISTANT") ~ "Assistant",
    pos("^SENIOR") ~ "Senior",
    pos("ADVISOR") ~ "Advisor",
    pos("^SPECIAL ASSISTANT") ~ "Special assistant",
    pos("^STAFF ASSISTANT") ~ "Staff assistant",
    TRUE ~ "Other")
        
ggplot(d, aes(x = Level, y = SALARY, color = ADMINISTRATION)) + 
    geom_boxplot() + 
    geom_point(position = position_dodge(width = 1)) + 
    gghlab::tilt_x_labels()

ggplot(d, aes(x = Level, fill = ADMINISTRATION)) + 
    geom_bar(stat = "count", position = "dodge") + 
    gghlab::tilt_x_labels()
```

## Gender

I decided to compare the genders of WH staffers and see if they got paid differently. To do this, I matched the first names of the staffers to a US database. If more than 75% of the people in the database with that name were one gender, I assigned that person that gender. If the name wasn't in the database or it was a more mixed-gender name (<75%), I looked the person up manually. This analysis unfortunately cannot account for transgender individuals as a separate category, so they get categorized according to the gender of their name.  

```{r}
d <- d %>% 
    mutate(FirstName = str_replace(NAME, "^.*, ", ""), 
        FirstName = str_replace_all(FirstName, " .*", ""))

library(gender)
genders <- gender(unique(d$FirstName), years = 1970)

# accept gender if proportion > 0.75
genders <- mutate(genders, gender = 
        ifelse(proportion_male < 0.75 & proportion_female < 0.75,
            "Indeterminate", gender), 
    FirstName = name)

# what names are indeterminate? 
# left_join(d, genders) %>% 
#     filter(is.na(gender) | gender == "Indeterminate") %>% 
#     select(NAME) %>% 
#     write.csv(file = "data/2017-week-29-names.csv", row.names = FALSE)

d <- left_join(d, genders)

manual_genders <- read.csv("data/2017-week-29-names.csv", 
    stringsAsFactors = FALSE) %>% 
    transmute(NAME = NAME, new_gender = gender)

to_plot <- left_join(d, manual_genders) %>% 
    mutate(gender = ifelse(new_gender %in% c("female", "male"), 
        new_gender, gender)) %>% 
    mutate(gender = ifelse(NAME == "Miraaj-Raza, Sidrah", "female", gender))

to_plot %>% filter(SALARY > 0) %>% 
    ggplot(aes(x = gender, y = SALARY)) + geom_boxplot() + geom_point() +
    facet_wrap(~ADMINISTRATION)
```

Overall, average salaries seem higher for men than women, especially in the Trump WH. Why?

```{r}
to_plot %>% filter(SALARY > 0) %>% 
    ggplot(aes(x = gender, y = SALARY)) + geom_boxplot() + geom_point() +
    facet_grid(ADMINISTRATION~Level)

to_plot %>% filter(SALARY > 0) %>% 
    ggplot(aes(x = gender, y = SALARY)) + geom_violin(draw_quantiles = 0.5) +
    geom_point() + 
    facet_grid(ADMINISTRATION~Level)

to_plot %>% filter(SALARY > 0) %>% 
    ggplot(aes(x = gender, y = SALARY)) + 
    ggbeeswarm::geom_quasirandom(method = "smiley", aes(color = gender)) +
    stat_summary(geom = "point", fun.y = mean) + 
    facet_grid(ADMINISTRATION~Level)

to_plot %>% filter(SALARY > 0, Level != "Chief") %>% 
    ggplot(aes(x = Level, y = SALARY, color = gender)) + 
    stat_summary(geom = "pointrange", fun.y = median, 
        fun.ymin = function(x) unname(quantile(x, 0)), 
        fun.ymax = function(x) unname(quantile(x, 1)), 
        position = position_dodge(width = 0.5)) + 
    facet_wrap(~ADMINISTRATION, ncol = 1)

to_plot %>% 
    mutate(title = str_sub(`POSITION TITLE`, 1, 20)) %>% 
    group_by(title, ADMINISTRATION, gender) %>%
    summarise(Salary = median(SALARY)) %>% 
    tidyr::spread(gender, Salary) %>% 
    filter(!is.na(male), !is.na(female)) %>% 
    group_by(title) %>% 
    filter(n() > 1) %>% 
    ggplot(aes(x = male, y = female)) + 
    geom_point() + facet_wrap(~ADMINISTRATION) + 
    geom_abline(slope = 1, intercept = c(0, 0))

to_plot %>% 
    mutate(title = str_sub(`POSITION TITLE`, 1, 20)) %>% 
    group_by(title, ADMINISTRATION, gender) %>%
    summarise(Salary = median(SALARY)) %>% 
    group_by(title) %>% 
    filter(n() > 3) %>% 
    arrange(desc(Salary)) %>% 
    ungroup() %>% 
    mutate(title = factor(title, levels = unique(title))) %>% 
    ggplot(aes(x = title, y = Salary, fill = gender)) + 
    geom_col(position = "dodge") + facet_wrap(~ADMINISTRATION) + 
    ggtitle("Median salary within a job level is\nsimilar across genders and admins")

ggplot(to_plot, aes(x = gender)) + 
    geom_bar(stat = "count") + facet_wrap(~ADMINISTRATION) + 
    ggtitle("Number of employees")

ggplot(to_plot, aes(x = SALARY, fill = gender)) + 
    geom_density(alpha = 0.5) + 
    facet_wrap(~ADMINISTRATION) + 
    ggtitle("Trump admin employs more men at higher-salary positions")

ggplot(to_plot, aes(x = SALARY, fill = gender)) + 
    geom_histogram(bins = 25) + 
    facet_grid(gender~ADMINISTRATION)

ggplot(to_plot, aes(x = SALARY, color = gender)) + 
    geom_freqpoly(bins = 10) + 
    facet_wrap(~ADMINISTRATION)

ggplot(to_plot, aes(x = SALARY, color = gender)) + 
    stat_ecdf(geom = "step") + facet_wrap(~ADMINISTRATION, ncol = 1) 
```

It looks to me like the difference in average salaries between men and women in the Trump admin are due to more men being employed in higher-wage positions, rather than men getting paid more than women who are in their same job category. 

```{r}
binned <- to_plot %>% 
    filter(SALARY > 0) %>%
    mutate(salary_bin = case_when(
        SALARY < 55000 ~ "<55K", 
        SALARY < 85000 ~ "55-85K",
        SALARY < 115000 ~ "85-115K", 
        SALARY < 145000 ~ "115-145K", 
        TRUE ~ ">145K"), 
        salary_bin = factor(salary_bin, levels = 
                c("<55K", "55-85K", "85-115K", "115-145K", ">145K")))

ggplot(binned, aes(x = salary_bin, fill = gender)) + 
    geom_bar(stat = "count", position = "dodge") + 
    facet_wrap(~ ADMINISTRATION, ncol = 1, scales = "free_y") + 
    gghlab::tilt_x_labels() + 
    scale_fill_manual(values = c("grey25", "grey50")) 

scaled_binned <- binned %>% 
    group_by(ADMINISTRATION, salary_bin, gender) %>%
    count() %>% 
    group_by(ADMINISTRATION, salary_bin) %>% 
    arrange(gender, .by_group = TRUE) %>% 
    summarise(MenPerWoman = n[2] / n[1])
    
ggplot(scaled_binned, aes(x = salary_bin, y = MenPerWoman, fill = ADMINISTRATION)) + 
    geom_col(position = "dodge") +  
    geom_hline(yintercept = 1) + 
    scale_fill_manual(values = c("#377eb8", "#e41a1c"), guide = FALSE) +
    ggtitle("Pay (dis)parity by gender in the White House") + 
    labs(y = "# of men employed per woman", 
        x = NULL, fill = NULL,
        subtitle = 
            paste0("More women work in low-wage positions. Higher-wage positions",
                "\nare dominated by men in the Trump White house.")) + 
    annotate("text", x = 0.4, y = c(2.85, 2.65), label = c("Obama", "Trump"), 
        color = c("#377eb8", "#e41a1c"), hjust = 0)

ggsave("out/2017-week-29.png", width = 5, height = 3.5, 
    units = "in", dpi = 303)        
```

Indeed, many more men than women in the higher wage positions than in the lower wage buckets. 